Elaboration with First-Class Implicit Function Types

### 2020-05-11 paper.tex
revs

\subsubsection{On metatheoretical properties of the core syntax}
We naturally want to make sure that the extended core syntax is still sensible.
In relation to type theories, the following three properties are commonly
considered.
\begin{itemize}
  \item A theory is \emph{consistent} if not every closed type has a closed
    inhabitant. An inconsistent theory cannot be used as a \emph{logic}, from
    the propositions-as-types perspective \cite[Chapter~1]{hottbook}, because
    there every proposition is provable, so proofs do not distinguish truth from
    falsehood.
  \item A theory has the \emph{canonicity} property, if every closed term is
    definitionally equal to a canonical term. For example, every closed term
    with a natural number type is a numeral, every $\mathsf{Bool}$ is $\true$ or
    $\false$ and so on. A canonical theory is useful as a \emph{total
      programming language}; if canonicity fails, then some programs compute
    to stuck terms which cannot be viewed as values. In non-total languages,
    progress and type preservation constitute a weaker analogue of canonicity.
  \item A theory has \emph{decidable conversion}, if definitional equality is
    decidable. Undecidable conversion usually implies undecidable type checking
    for type theories. Often, decidable conversion follows from a normalization
    procedure for open terms: we first normalize, then decide equality of normal
    forms.
\end{itemize}

@@

In any case, a \emph{reduction} of new features to old features is a common and
convenient strategy to reduce metatheoretical properties of extended theories to
that of smaller or more conventional theories. In the simplest case, new
features are definable from old features. Here, we can either assume that new
features are merely definitional shorthands, or give an evident injective
translation from the new theory to the old one. Both imply that the desired
properties transfer from the old theory to the new one.

@@

\begin{figure}
\begin{alignat*}{4}
  & \mi{IdTy}   && :\equiv \{A : \U\}\to A \to A   \hspace{10em}         && \mi{inc}    && : \mi{Int} \to \mi{Int}                                        \\
  & \mi{single} && : \{A : \U\} \to A \to \mi{List}\,A                        && \mi{auto}   && : IdTy \to IdTy                                      \\
  & \mi{id}     && : IdTy                                                && \mi{auto'}  && : \{B : \U\}\to IdTy \to B \to B                     \\
  & \mi{ids}    && : \mi{List}\,IdTy                                          && \mi{choose} && : \{A : \U\} \to A \to A \to A                       \\
  & \mi{nil}    && : \{A : \U\}\to \mi{List}\,A                               && \mi{app}    && : \{A\,B : \U\} \to (A \to B) \to A \to B            \\
  & \mi{cons}   && : \{A : \U\} \to A \to \mi{List}\,A \to \mi{List}\,A            && \mi{revapp} && : \{A\,B : \U\} \to A \to (A \to B) \to B            \\
  & \mi{head}   && : \{A : \U\} \to \mi{List}\,A \to A                        && \mi{runST}  && : \{A : \U\} \to (\{S : \U\} \to \mi{ST}\,S\,A) \to A     \\
  & \mi{tail}   && : \{A : \U\} \to \mi{List}\,A \to \mi{List}\,A                  && \mi{argST}  && : \{S : \U\}\to \mi{ST}\,S\,\mi{Int}                           \\
  & \mi{map}    && : \{A\,B : \U\} \to (A \to B) \to \mi{List}\,A \to \mi{List}\,B && \mi{poly}   && : IdTy \to \mi{Pair}\,\mi{Int}\,\mi{Bool}
\end{alignat*}
\caption{Types used in Figure \ref{fig:benchmark}.}
\label{fig:benchmark_types}
\end{figure}

@@

FCIF cannot infer the $B1$ and $B2$ cases on Figure \ref{fig:benchmark}, which
would require inferring polymorphic types for function arguments. In very simple
cases, FCIF is able to infer (ambiguous) polymorphic argument types. For
example, it can infer a type for $(\lambda\,f.\,f\{\mi{zero}\})$. Assuming
$\mi{Pair} : \U \to \U \to \U$ and $\mi{pair}: \{A\,B : \U\}\to A \to B \to
\mi{Pair}\,A\,B$, it can also infer type for the following:
\[
    \lambda\,f\,x.\,\mi{pair}\,(f\,\{x\}\,x)\,(f\,x)
\]
Here, FCIF can proceed because it first encounters $(f\,\{x\}\,x)$, where it
learns that $f$ should have two arguments, where the first one is implicit.
Armed with this knowledge, it inserts a fresh meta in $(f\,x)$. The following
fails though:
\[
    \lambda\,f\,x.\,\mi{pair}\,(f\,x)\,(f\,\{x\}\,x)
\]
Now, FCIF decides upon elaborating $(f\,x)$ that $f$ is a function whose first
argument is explicit, and fails at $f\,\{x\}$. This is similar to the implicit
$\lambda$ insertion issue, but here we would have to refine implicit
applications. We could try inserting an unknown curried application after the
first $f$, which could be later refined. We sketch an extension of elaboration
which could possibly handle this example. We extend insertion with
the case which inserts a new curried application whenever we have a meta-headed
type. We omit metacontexts and metasubstitutions for brevity.
\begin{alignat*}{3}
& \einsert\,(t,\,\alpha\,\ol{u}) :\equiv \edo \\
& \quad A \leftarrow \mathsf{freshMeta}_{\Gamma}\,\Tel\\
& \quad B \leftarrow \mathsf{freshMeta}_{\Gamma,\,x: \Rec\,A}\,\U\\
& \quad \unify\,(\alpha\,\ol{u})\,(\{x : \ol{A}\} \to B)\\
& \quad \mathsf{newConstancy}_{\Gamma, x : \Rec\,A}\,B\\
& \quad u \leftarrow \mathsf{freshMeta}_{\Gamma}\,(\Rec\,A)\\
& \quad \ereturn\,(t\,\{u : \ol{A}\},\,B[x \mapsto u])
\end{alignat*}

%% let e2 = λ f (x : Int). pair (f {id {Int} x} x) (f x) in

It is apparent from the $B1$ and $B2$ cases on Figure \ref{fig:benchmark} that FCIF
cannot infer implicit function types for function arguments.  MLF is the only
system which can do this, in cases where the polymorphic argument is used only
once. Consider the following:
\[
  \slet\,f : \mi{IdTy} \to \mi{Bool} = \lambda\,\_.\,\mi{true}\,\sin\,\lambda\,x.\,f\,x
  \]
FCIF will attempt to elaborate $f\,x$ to $f\,(\lambda\{A\}.\,x)$, but fails to
infer a type for $x$, because $A$ is not in the scope of $x$'s type. We sketch
an extension of FCIF with \emph{curried application insertion}, which could
possibly handle this example. We extend insertion with the following case. We
omit metacontexts and metasubstitutions for brevity.
\begin{alignat*}{3}
& \einsert\,(t,\,\alpha\,\ol{u}) :\equiv \edo \\
& \quad A \leftarrow \mathsf{freshMeta}_{\Gamma}\,\Tel\\
& \quad B \leftarrow \mathsf{freshMeta}_{\Gamma,\,x: \Rec\,A}\,\U\\
& \quad \unify\,(\alpha\,\ol{u})\,(\{x : \ol{A}\} \to B)\\
& \quad \mathsf{newConstancy}_{\Gamma, x : \Rec\,A}\,B\\
& \quad u \leftarrow \mathsf{freshMeta}_{\Gamma}\,(\Rec\,A)\\
& \quad \ereturn\,(t\,\{u : \ol{A}\},\,B[x \mapsto u])
\end{alignat*}
In short, we insert a new curried application whenever we $\einsert$ with
meta-headed type. Now, elaboration returns
$f\,(\lambda\,\{A\}.\,x\,\{\alpha_2\,x\,A : \ol{\alpha_0}\})$, and right before
we try to unify expected and inferred types, we have $x : \{y :
\ol{\alpha_0}\}\to \alpha_1\,\{y : \ol{\alpha_0}\}$, and we have to compute
$\unify\,(\alpha_1\,\{\alpha_2\,x\,A : \ol{\alpha_0}\})\,(A \to A)$. This
problem does not fall in any usual pattern fragment, and does not have an
obvious most general solution. However, it is reasonable to make the following
assumption: metas which return in unknown record types, like $\alpha_2$, are
only solved with \emph{order-preserving embeddings}, so for $\alpha_2$ the four
potential solutions are $(\lambda\,x\,A.\,[])$, $(\lambda\,x\,A.\,(A :: []))$,
$(\lambda\,x\,A.\,(x :: []))$ and $(\lambda\,x\,A.\,(x :: A :: []))$\footnote{Technically, we could also allow permutations such as $\lambda\,x\,A.\,(A :: x :: [])$, but these are superfluous since nothing in our system depends on the ordering of permutable function arguments.}. The reason
for only allowing such solutions is that these can possibly yield solvable
pattern unification problems, while solutions containing arbitrary terms or
non-linear variable occurrences cannot. Out of these, $(\lambda\,x\,A.\,(x :: A
:: []))$ and $(\lambda\,x\,A.\,(x :: []))$ are ruled out because they would
require a cyclic solution for $\alpha_0$, and $(\lambda\,x\,A.\,[])$ is ruled
out because it would yield the unsolvable $\unify\,\alpha_1\,(A \to A)$
problem. Thus, $(\lambda\,x\,A.\,(A :: []))$ is the unique order-preserving
embedding solution in this case, which does yield the expected elaboration
output. However, this idea is far from being fully fleshed out, and it
could be subject of future research.

>>>>

\subsubsection{On metatheoretical properties of the core syntax}
We naturally want to make sure that the extended core syntax is still sensible.
In relation to type theories, the following three properties are commonly
considered.
\begin{itemize}
  \item A theory is \emph{consistent} if not every closed type has a closed
    inhabitant. An inconsistent theory cannot be used as a \emph{logic}, from
    the propositions-as-types perspective \cite[Chapter~1]{hottbook}, because
    there every proposition is provable, so proofs do not distinguish truth from
    falsehood.
  \item A theory has the \emph{canonicity} property, if every closed term is
    definitionally equal to a canonical term. For example, every closed natural
    number is a numeral, every $\mathsf{Bool}$ is $\true$ or $\false$ and so
    on. A canonical theory is useful as a \emph{total programming language}; if
    canonicity fails, then some programs compute to stuck terms which cannot be
    viewed as values. In non-total languages, progress and type preservation
    constitute a weaker analogue of canonicity.
  \item A theory has \emph{decidable conversion}, if definitional equality is
    decidable. Undecidable conversion usually implies undecidable type checking
    for type theories. Often, decidable conversion follows from a normalization
    procedure for open terms: we first normalize, then decide equality of normal
    forms.
\end{itemize}

@@

In any case, a \emph{reduction} of new features to old features is a common and
convenient strategy to reduce metatheoretical properties of extended theories to
that of smaller or better-known theories. See \citet{next700} for a collection
of examples for this. In the simplest case, new features are definable from old
features. Here, we can either assume that new features are merely definitional
shorthands, or give an evident injective translation from the new theory to the
old one. Both imply that the desired properties transfer from the old theory to
the new one.

@@

\begin{figure}
\begin{alignat*}{4}
  & \mi{IdTy}   && :\equiv \{A : \U\}\to A \to A   \hspace{10em}         && \mi{inc}    && : \mi{Int} \to \mi{Int}                                        \\
  & \mi{single} && : \{A : \U\} \to A \to \mi{List}\,A                        && \mi{auto}   && : IdTy \to IdTy                                      \\
  & \mi{id}     && : IdTy                                                && \mi{auto'}  && : \{B : \U\}\to IdTy \to B \to B                     \\
  & \mi{ids}    && : \mi{List}\,IdTy                                          && \mi{choose} && : \{A : \U\} \to A \to A \to A                       \\
  & \mi{nil}    && : \{A : \U\}\to \mi{List}\,A                               && \mi{app}    && : \{A\,B : \U\} \to (A \to B) \to A \to B            \\
  & \mi{cons}   && : \{A : \U\} \to A \to \mi{List}\,A \to \mi{List}\,A            && \mi{revapp} && : \{A\,B : \U\} \to A \to (A \to B) \to B            \\
  & \mi{head}   && : \{A : \U\} \to \mi{List}\,A \to A                        && \mi{runST}  && : \{A : \U\} \to (\{S : \U\} \to \mi{ST}\,\,S\,A) \to A     \\
  & \mi{tail}   && : \{A : \U\} \to \mi{List}\,A \to \mi{List}\,A                  && \mi{argST}  && : \{S : \U\}\to \mi{ST}\,\,S\,\mi{Int}                           \\
  & \mi{map}    && : \{A\,B : \U\} \to (A \to B) \to \mi{List}\,A \to \mi{List}\,B && \mi{poly}   && : IdTy \to \mi{Pair}\,\mi{Int}\,\mi{Bool}
\end{alignat*}
\caption{Types used in Figure \ref{fig:benchmark}.}
\label{fig:benchmark_types}
\end{figure}

@@

FCIF cannot infer the $B1$ and $B2$ cases on Figure \ref{fig:benchmark}, which
would require inferring polymorphic types for function arguments. In very simple
cases, FCIF is able to infer (ambiguous) polymorphic argument types. For
example, it can infer a type for $(\lambda\,f.\,f\{\mi{zero}\})$. Assuming
$\mi{Pair} : \U \to \U \to \U$ and $\mi{pair}: \{A\,B : \U\}\to A \to B \to
\mi{Pair}\,A\,B$, it can also infer type for the following:
\[
    \lambda\,f\,x.\,\mi{pair}\,(f\,\{x\}\,x)\,(f\,x)
\]
Here, FCIF can proceed because it first encounters $(f\,\{x\}\,x)$, where it
learns that $f$ should have two arguments, where the first one is implicit.
Armed with this knowledge, it inserts a fresh meta in $(f\,x)$. The following
fails though:
\[
    \lambda\,f\,x.\,\mi{pair}\,(f\,x)\,(f\,\{x\}\,x)
\]
Here, FCIF decides upon elaborating $(f\,x)$ that $f$ is a function whose first
argument is explicit, and fails at $f\,\{x\}$. This is similar to the implicit
$\lambda$ insertion issue, but here we would have to refine implicit
applications. We could try inserting an unknown curried application after the
first $f$, which could be later refined. Below we sketch an extension of
$\einsert$ which accomplishes this. We omit metacontexts and metasubstitutions
for brevity.
\begin{alignat*}{3}
& \einsert\,(t,\,\alpha\,\ol{u}) :\equiv \edo \\
& \quad A \leftarrow \mathsf{freshMeta}_{\Gamma}\,\Tel\\
& \quad B \leftarrow \mathsf{freshMeta}_{\Gamma,\,x: \Rec\,A}\,\U\\
& \quad \unify\,(\alpha\,\ol{u})\,(\{x : \ol{A}\} \to B)\\
& \quad \mathsf{newConstancy}_{\Gamma, x : \Rec\,A}\,B\\
& \quad u \leftarrow \mathsf{freshMeta}_{\Gamma}\,(\Rec\,A)\\
& \quad \ereturn\,(t\,\{u : \ol{A}\},\,B[x \mapsto u])
\end{alignat*}

However, this alone does not work: in $t\,\{u : \ol{A}\}$, we return a term
applied to a record with \emph{unknown telescope type}, which later may give
rise to unification problems where metas are applied to such unknown telescopes.
Our current unification algorithm does not handle these cases meaningfully, and
we are not aware of any applicable algorithm in prior literature.

A robust system for this would give a solution for polymorphic argument
inference. We do not expect that this is easy. The only prior work which
supports this feature is MLF, and recall that MLF's inference algorithms are
already very complex, and MLF does not support dependent types.

We nevertheless sketch a solution, which is not yet fleshed out, but could be
developed in future work. We work with the following example:
\[
  \slet\,f : \mi{IdTy} \to \mi{Bool} = \lambda\,\_.\,\mi{true}\,\sin\,\lambda\,x.\,f\,x
\]
Plain FCIF will attempt to elaborate $f\,x$ to $f\,(\lambda\{A\}.\,x)$, but
fails to infer a type for $x$, because $A$ is not in the scope of $x$'s type.
FCIF with curried application insertion instead returns
$f\,(\lambda\,\{A\}.\,x\,\{\alpha_2\,x\,A : \ol{\alpha_0}\})$ for the same
expression, and we also have $x : \{y : \ol{\alpha_0}\}\to \alpha_1\,\{y :
\ol{\alpha_0}\}$. However, at this point FCIF still needs to unify expected and
inferred types, so it needs to compute
\[
\unify\,(\alpha_1\,\{\alpha_2\,x\,A : \ol{\alpha_0}\})\,(A \to A)
\]
This does not fall into any pattern fragment, and has no most general solution.
However, we may make several assumptions about unknown telescopes in spines. We
can view this as an extension of Miller's pattern unification
\cite{millerpattern} with unknown spine slices: $\alpha_2\,x\,A$ stands for a
part of the spine which may contain bound variables $x$ and $A$. We make the
following assumption: $\alpha_2$ may be only solved with \emph{order-preserving
  embeddings}, so the four potential solutions are $(\lambda\,x\,A.\,[])$,
$(\lambda\,x\,A.\,(A :: []))$, $(\lambda\,x\,A.\,(x :: []))$ and
$(\lambda\,x\,A.\,(x :: A :: []))$. On the one hand, this rules out solutions
which contain arbitrary terms or non-linear variable occurrences, which would
yield unsolvable patterns. On the other hand, we probably do not throw away many
useful solutions, because all curried applications arise from elaboration, and
we do not have to deal with any of them coming from the surface language. In
short, we try to keep the invariant that metas returning in record types are
only ever solved with order-preserving embeddings.

Out of the four possible embeddings, $(\lambda\,x\,A.\,(x :: A :: []))$ and
$(\lambda\,x\,A.\,(x :: []))$ are ruled out because they would require a cyclic
solution for $\alpha_0$, and $(\lambda\,x\,A.\,[])$ is ruled out because it
would yield the unsolvable $\unify\,\alpha_1\,(A \to A)$ problem. Thus,
$(\lambda\,x\,A.\,(A :: []))$ is the unique order-preserving embedding solution
in this case, which does yield the expected elaboration output. We note though
that ``unique order-preserving embedding'' does not yet provide an efficient
algorithmic implementation, and we would need to be smarter in considering
particular unification problems, to avoid backtracking through possible
solutions.

%% It is apparent from the $B1$ and $B2$ cases on Figure \ref{fig:benchmark} that FCIF
%% cannot infer implicit function types for function arguments.  MLF is the only
%% system which can do this, in cases where the polymorphic argument is used only
%% once. Consider the following:

%% FCIF will attempt to elaborate $f\,x$ to $f\,(\lambda\{A\}.\,x)$, but fails to
%% infer a type for $x$, because $A$ is not in the scope of $x$'s type. We sketch
%% an extension of FCIF with \emph{curried application insertion}, which could
%% possibly handle this example. We extend insertion with the following case. We
%% omit metacontexts and metasubstitutions for brevity.
%% \begin{alignat*}{3}
%% & \einsert\,(t,\,\alpha\,\ol{u}) :\equiv \edo \\
%% & \quad A \leftarrow \mathsf{freshMeta}_{\Gamma}\,\Tel\\
%% & \quad B \leftarrow \mathsf{freshMeta}_{\Gamma,\,x: \Rec\,A}\,\U\\
%% & \quad \unify\,(\alpha\,\ol{u})\,(\{x : \ol{A}\} \to B)\\
%% & \quad \mathsf{newConstancy}_{\Gamma, x : \Rec\,A}\,B\\
%% & \quad u \leftarrow \mathsf{freshMeta}_{\Gamma}\,(\Rec\,A)\\
%% & \quad \ereturn\,(t\,\{u : \ol{A}\},\,B[x \mapsto u])
%% \end{alignat*}
%% In short, we insert a new curried application whenever we $\einsert$ with
%% meta-headed type. Now, elaboration returns
%% $f\,(\lambda\,\{A\}.\,x\,\{\alpha_2\,x\,A : \ol{\alpha_0}\})$, and right before
%% we try to unify expected and inferred types, we have $x : \{y :
%% \ol{\alpha_0}\}\to \alpha_1\,\{y : \ol{\alpha_0}\}$, and we have to compute
%% $\unify\,(\alpha_1\,\{\alpha_2\,x\,A : \ol{\alpha_0}\})\,(A \to A)$. This
%% problem does not fall in any usual pattern fragment, and does not have an
%% obvious most general solution. However, it is reasonable to make the following
%% assumption: metas which return in unknown record types, like $\alpha_2$, are
%% only solved with \emph{order-preserving embeddings}, so for $\alpha_2$ the four
%% potential solutions are $(\lambda\,x\,A.\,[])$, $(\lambda\,x\,A.\,(A :: []))$,
%% $(\lambda\,x\,A.\,(x :: []))$ and $(\lambda\,x\,A.\,(x :: A :: []))$\footnote{Technically, we could also allow permutations such as $\lambda\,x\,A.\,(A :: x :: [])$, but these are superfluous since nothing in our system depends on the ordering of permutable function arguments.}. The reason
%% for only allowing such solutions is that these can possibly yield solvable
%% pattern unification problems, while solutions containing arbitrary terms or
%% non-linear variable occurrences cannot. Out of these, $(\lambda\,x\,A.\,(x :: A
%% :: []))$ and $(\lambda\,x\,A.\,(x :: []))$ are ruled out because they would
%% require a cyclic solution for $\alpha_0$, and $(\lambda\,x\,A.\,[])$ is ruled
%% out because it would yield the unsolvable $\unify\,\alpha_1\,(A \to A)$
%% problem. Thus, $(\lambda\,x\,A.\,(A :: []))$ is the unique order-preserving
%% embedding solution in this case, which does yield the expected elaboration
%% output. However, this idea is far from being fully fleshed out, and it
%% could be subject of future research.
